Columnar VCF for Biobank scale datasets using Zarr

### 2024-04-11 paper.tex
outline of GeL narrative.
Closes #4

\subsection{Case study: Genomics England 100,000 genomes}
In this section we demonstrate the utility of vcfzarr on a large human dataset.
Genomics England’s multi-sample VCF dataset (aggV2) is an 
aggregate of 78,195 gVCFs from rare disease and cancer participants 
recruited as part of the 100,000 Genomes Project~\cite{turnbull2018100}. 
The dataset comprises approximately 722 million annotated single-nucleotide 
variants (SNVs) and small indels split into 1,371 roughly equal chunks and 
totalling 165.3 TiB of VCF data after \texttt{bgzip} compression. 
The dataset is used for a variety of research purposes, ranging from 
genome-wide association studies (GWAS)~\cite{kousathanas2022whole} and 
imputation~\cite{shi2023genomics} to 
simple queries involving single gene 
regions~\cite{leggatt2023genotype,lam2023repeat}.

We converted the data to the intermediate columnar format 
which took X hours with a cluster of Y workers using the 
distributed explode commands. When then converted to Zarr using 
SOMETHING which took X hours using [RESOURCES]. In total, this gave 
X TiB of data, spread across Y thousand files, a compression of 
X over the original VCFs.
The detailed breakdown of per-column compression ratios is 
shown in Table~\ref{tab-genomics-england-data}.

\begin{table}
\caption{Summary of vcf-zarr performance on the Genomics England 100,000
genomes VCF dataset.
\label{tab-genomics-england-data}}
\begin{tabular}{llS[table-format=3.1]S[table-format=3.2]S[table-format=3.2]}
\toprule
{Field} & {type} & {storage} & {compress} & {\%total} \\
\midrule
call\_AD &  int16 & 179.0G & 26 & 24.1\\
call\_GQ &  int16 & 171.0G & 13 & 22.2 \\
call\_DP &  int16 & 141.0G & 20 & 18.8 \\
call\_FT &  string & 58.0G & 160 & 7.3 \\
call\_PL &  int16 & 51.0G & 140 & 6.8 \\
call\_GQX &  int16 & 12.0G & 190 & 1.6 \\
call\_genotype & int8 & 6.1G & 380 & 0.8 \\
call\_genotype\_mask & bool & 3.6G  & 630 & 0.48\\
call\_genotype\_phased & bool & 3.6G  & 1700 & 0.48 \\
variant\_quality & float32 & 22.6M & 2.7 & <0.01 \\
variant\_allele & string & 21.5M & 11 \\
variant\_position & int32 & 12.7M & 4.7 \\
variant\_ABratio & float32 & 9.6M & 6.3 \\
variant\_AN & int32 & 9.6M & 6.3 \\
variant\_filter & bool & 1.9M & 490 \\
\bottomrule
\end{tabular}
\end{table}

Table~\ref{tab-genomics-england-data} shows that the dataset size is 
dominated by a few columns. This is well known and has been remarked 
on several times [e.g. spVCF paper, etc]. [In this paragraph we discuss
how some of these fields can be considerably compressed with very
little loss of information. For example, the maximum Allele Depth (AD)
value seen in this dataset is >3000. However, when we compute the 
histogram of AD values we can see that something like 0.1\% of 
AD values are > 127, so we can reasonably store these as 8 bit 
ints. This reduces storage to X. Finish up by saying that more 
could be done, but it's beyond the scope of this paper. Maybe note 
that this kind of per-field analysis is easy now, e.g. computing 
the hist took X minutes using xarray and Dask on a cluster with
Y workers..]

To illustrate the difference in query performance we ran some 
\texttt{bcftools query} commands based on the Genomics England 
user guides. For example, finding all variants passing 
quality filters with an allele frequency X takes just 
x seconds on a single core (y seconds of which is writing the 
CSV file). The same query took X hours on one chunk, giving a 
an approximate total time of XX hours over all Y chunks 
in chromosomes 2.

>>>>

\subsection{Case study: Genomics England 100,000 genomes}
In this section we demonstrate the utility of \texttt{vcfzarr} on a large human dataset
and the scalability of the \texttt{vcf2zarr} conversion utility.
Genomics England’s multi-sample VCF dataset (aggV2) is an 
aggregate of 78,195 gVCFs from rare disease and cancer participants 
recruited as part of the 100,000 Genomes Project~\cite{turnbull2018100}. 
The dataset comprises approximately 722 million annotated single-nucleotide 
variants (SNVs) and small indels split into 1,371 roughly equal chunks and 
totalling 165.3 TiB of VCF data after \texttt{bgzip} compression. 
The dataset is used for a variety of research purposes, ranging from 
genome-wide association studies (GWAS)~\cite{kousathanas2022whole} and 
imputation~\cite{shi2023genomics} to 
simple queries involving single gene 
regions~\cite{leggatt2023genotype,lam2023repeat}.

As described in the Methods, conversion to Zarr using 
\texttt{vcf2zarr} is a two-step 
process to limit memory usage and ensure reliability. We 
first convert the X VCF files (YTiB) for chromosome 2
into the intermediate columnar format. This task was 
distributed over a cluster of N workers, using a 
total of X CPU hours and an elapsed wall-clock time 
of X hours. When then converted the intermediate 
column format to Zarr using 
SOMETHING which took X hours using [RESOURCES]. 
This produced a dataset with N arrays, consuming a 
total of Y TiB of storage over X directories and 
Y files. The top fields in terms 
of storage are detailed in Table~\ref{tab-genomics-england-data}.

\begin{table}
\caption{Summary of the top N vcfzarr columns produced for 
Genomics England aggV2 VCFs on chromosome 20 using \texttt{vcf2zarr}
default settings. Each field is stored independently 
as Zarr array with the given type (sufficient to represent all values in the
data). We show the total storage consumed (reported via \texttt{du}) in 
power-of-two units, and the compression ratio achieved on that array.
We also show the percentage of the overall storage that each array consumes.
\label{tab-genomics-england-data}}
\begin{tabular}{llS[table-format=3.1]S[table-format=3.2]S[table-format=3.2]}
\toprule
{Field} & {type} & {storage} & {compress} & {\%total} \\
\midrule
call\_AD &  int16 & 179.0G & 26 & 24.1\\
call\_GQ &  int16 & 171.0G & 13 & 22.2 \\
call\_DP &  int16 & 141.0G & 20 & 18.8 \\
call\_FT &  string & 58.0G & 160 & 7.3 \\
call\_PL &  int16 & 51.0G & 140 & 6.8 \\
call\_GQX &  int16 & 12.0G & 190 & 1.6 \\
call\_genotype & int8 & 6.1G & 380 & 0.8 \\
call\_genotype\_mask & bool & 3.6G  & 630 & 0.48\\
call\_genotype\_phased & bool & 3.6G  & 1700 & 0.48 \\
variant\_quality & float32 & 22.6M & 2.7 & <0.01 \\
variant\_allele & string & 21.5M & 11 \\
variant\_position & int32 & 12.7M & 4.7 \\
variant\_ABratio & float32 & 9.6M & 6.3 \\
variant\_AN & int32 & 9.6M & 6.3 \\
variant\_filter & bool & 1.9M & 490 \\
\bottomrule
\end{tabular}
\end{table}

Table~\ref{tab-genomics-england-data} shows that the dataset storage
size dominated by a few columns, with call\_AD and call\_GQ consuming 
almost half of total. These fields are much less compressible
than genotype data (which uses $<1\%$ of the total space here)
due to their inherent noisiness~\citep{lin2020sparse}. Substantial
reductions in space can be achieved, however, if we examine the 
distributions of these fields and truncate some outliers.
Figure~SX shows the histogram of the XX call\_AD values, and we can
see that < 0.0X\% of the values are $>127$. Truncating and 
re-encoding the call\_AD array as an int8 results in a total storage of 
XX. A substantial advantage of Zarr is that such per-field optimisations
can be made after conversion, without affecting the rest of the dataset.
As discussed in Methods, judicious use of additional filters
and lossy quantizing can result substantial space savings.

Calculating the histogram of call\_AD values at this 
scale is not a trivial task.
Taking advantage of the rich software ecosystem growing around Zarr
(see Discussion) we used xarray~\citep{hoyer2017xarray}
and Dask~\citep{rocklin2015dask} to transparently distribute the 
computation over a cluster of X workers. The elapsed time 
was X minutes, and requires X lines of Python in a 
Jupyter notebook~\citep{kluyver2016jupyter}.

One of the key uses of the aggV2 dataset is to enable simple queries 
of specific regions. For example, in [A PAPER] they [DID SOMETHING].
We replicated this by running the appropriate bcftools query
on the VCF chunk (note that the user is responsible for finding 
the correct chunk file here, limiting the utility of genomic 
indexes). This took XXX hours.
The same query implemented with the Zarr-python 
API required XXX seconds to produce identical output.
